---
title: "Summer of Tech NetworkD3 plots Dec 2017"
author: "Kim Fitter"
date: "27 June 2018"
output: 
      html_document
---

This R Markdown plots a Sankey diagram and a force directed network using [networkD3](https://cran.r-project.org/web/packages/networkD3/index.html).

```{r setup, echo=FALSE, results='asis'}
knitr::opts_chunk$set(warning=FALSE)
```


```{r load packages, message=FALSE, warning=FALSE}
library(tidyverse)
library(networkD3)
```

## Data 

The Summer of Tech data is available in the R-Ladies github [link](https://github.com/R-LadiesAKL/sotdata). 

```{r import data, message=FALSE}
# Import the "Event attendances data Dec 2017.csv"
event <- read_csv("https://raw.githubusercontent.com/R-LadiesAKL/sotdata/master/Event%20attendances%20data%20Dec%202017.csv")
# Import the intern data "Intern data Dec 2017.csv"
intern <- read_csv("https://raw.githubusercontent.com/R-LadiesAKL/sotdata/master/Intern%20data%20Dec%202017.csv")
#  Only select the first 214 rows, the rest are N/A
intern <- intern[1:214,]
```
  


```{r make names}
#  We will use the make.names base R function to make syntactically valid names of the column variables for further analysis and output in R Markdown. For these names, it adds a "." to the spaces in the names
#  Make Syntactically Valid Names of the event
names(event) <- make.names(names(event))
#  Make Syntactically Valid Names of the intern
names(intern) <- make.names(names(intern))
```



## Clean Data

**Duplicates**

From the summary statistics there appears to be a student who has two internships? Let's view and remove the duplicate token.

```{r duplicate internships}
#  Check which internship with the duplicated token ID
intern %>% 
      filter(token==intern$token[duplicated(intern$token)]) %>% 
      kableExtra::kable() %>%
      kableExtra::kable_styling()
# Remove the duplicate token by identifying the row number with the duplicate and then using dplyr's slice function
intern <- intern %>% 
      slice(-which(duplicated(intern$token)))
```

**Add successful intern flag**

Add a variable to the intern data to identify successful internship students after joining with the event data.

```{r clean files}
# Add a variable to the intern data to identify successful internship students after merged with the event data
intern <- intern %>% 
      mutate(internship="Successful Internship")
```

**Rename column variables**

Let's rename some of the column variables to be able to track which dataset set they came from.

```{r rename cols}
# Rename column variables
intern <- intern %>% 
      # Rename the intern Region to intern.region
      rename_at(vars('Region'), ~ 'Intern.region') %>% 
      # Rename the intern institution to intern.institution
      rename_at(vars('Institution'), ~ 'Intern.institution') %>% 
      # Rename the intern Year of study to Intern.year.of.study
      rename_at(vars('Year.of.study'), ~ 'Intern.year.of.study')
event <- event %>% 
       # Rename the Student token variable in event to token so that this is the same as the intern variable name.
      rename_at(vars('Student.token'), ~ 'token') %>% 
      # Rename the region institution to region.institution
      rename_at(vars('Institution'), ~ 'Event.institution') 
```
  


**Join files**

Next Join the intern and event files by the common identifier, the student ID, "token".

```{r join}
# Join the intern and event by the common token variable
combine <- full_join(intern,event,by="token")
# Replace the internship NA's to no. This will include event attendance states including late withdrawals
combine$internship <- replace_na(combine$internship,"Unsuccessful Internships")
# Since some students attend multiple events, let's create a flag for attended events
combine <- combine %>% 
      mutate(registered.for.events=ifelse(is.na(Event.region),"no","yes"))
# Create a new student institution variable from the Event and intern institution variables
combine <- combine %>% 
      mutate(Institution=ifelse(is.na(Event.institution),Intern.institution, Event.institution))
```

To view success factors I  wanted to visualise the flows from one set of values to another i.e. flow from a factor to whether the student was successful or unsuccessful with an internship.

**Institutions**

The flow I decided to look at is  the flow of number of students from institutions to successful and unsuccessful internships.

I initially tried the Use the gvisSankey function from the [googleVis](https://cran.r-project.org/web/packages/googleVis/) R package. The downside is that it creates a plot in a new browser tab and not easily embedded to R Markdown so not suitable for use here.

Then I tried the [networkD3](https://cran.r-project.org/web/packages/networkD3/index.html) R package to create a D3 JavaScript Sankey diagram.

I included a check that the total number of intern tokens in original file is same as sum in this summary after the join and also include a new "Student Institution" variable.

See below the interactive Sankey diagram:

```{r instition to yes or no internships, results='asis'}
# Create a summary of the counts of Event institutions by unique interns
inst_intern_count <- combine %>%
      distinct(Event.institution,internship,token) %>%
      group_by(Event.institution,internship) %>%  
      summarize(counts = n()) %>%
      ungroup() %>%  
      arrange(desc(counts))
# Replace the Event NA's to Other
inst_intern_count$Event.institution <- replace_na(inst_intern_count$Event.institution,"Other Institutions")
# Since there is a large number of institutions, Select some to show and group the rest as "Other"
keep <- c("Victoria University of Wellington","Auckland University","Auckland University of Technology","Whitireia New Zealand","Wellington Institute of Technology (Weltec)","Massey University (Wellington)")
notkeep <- setdiff(unique(inst_intern_count$Event.institution), keep)
inst_intern_count$Event.institution[inst_intern_count$Event.institution %in% notkeep] <- "Other"
# Check that the total intern tokens in original file is same as sum in this summary
length(unique(intern$token))==
inst_intern_count %>% filter(internship=="Successful Internship") %>% select(counts) %>% sum()

# Create the links and nodes for the Sankey and force network diagrams
name_vec <- c(unique(inst_intern_count$Event.institution), unique(inst_intern_count$internship))
name_vec <- name_vec[!is.na(name_vec)]
lengthid <- length(name_vec)
nodes <- data.frame(name = name_vec,id=0:(lengthid-1),stringsAsFactors = FALSE)
links <- inst_intern_count %>%
  left_join(nodes, by = c('Event.institution' = 'name')) %>%
  rename(origin_id = id) %>%
  left_join(nodes, by = c('internship' = 'name')) %>%
  rename(dest_id = id) %>% 
      data.frame()
```


```{r sankey plot, results='asis'}
sankeyNetwork(Links = links, Nodes = nodes, Source = 'origin_id', Target = 'dest_id', Value = 'counts', NodeID = 'name', 
              # Format
              fontSize = 14,
              nodeWidth = 20,
              fontFamily = "sans-serif")   

# Use this to save the plot as html
# saveNetwork(file = 'sankey.html')
```

Let's view this with an interactive and zoom-able force directed network graph using the forceNetwork function from the same [networkD3](https://cran.r-project.org/web/packages/networkD3/index.html) R package. 

```{r force network, results='asis'}
# Create and plot the forcenetwork with the links and nodes
forceNetwork(Links = links, Nodes = nodes, Source = 'origin_id', Target = 'dest_id', NodeID = 'name', Group = 'id', 
             # Customise the edges
             Value = 'counts', 
             # Layout
             #height = 500, 
             #width = 500, 
             zoom = TRUE,
             # Customise the links
             linkWidth = networkD3::JS("function(d) { return d.value/50; }"),
             charge=-5000,
             # Format
             fontFamily = "sans-serif",
             opacity = 0.95, 
             opacityNoHover = 0.5,
             fontSize = 18)   

# Use this to save the plot as html
# saveNetwork(file =  'forcediag.html',selfcontained = TRUE)
```
  


